Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Introduce treatments / nnt before nns

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).

@@

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
The number needed to treat (NNT) is the number of true positive patients that
must be treated by an intervention in order for one patient to benefit from the
treatment. 
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
The NNB pairs model performance with treatment effectiveness to estimate the
benefit of using predictive models in clinical practice.

We determined the decision threshold at the 95th percentile of risk for each
prediction model and computed the confusion matrix and several performance
metrics including the NNS (@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than prior curated models and it is the most clinically
relevant as physician chart review increases confidence that positively-labelled
severe outcomes are due to the CDI rather than other causes.

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
sustained resolution of CDI and reduced risk of recurrence
[@stuart_clinical_2021].
However, fidaxomicin is considerably more expensive than vancomycin, resulting
in most CDI patients in the US being treated with vancomycin (TODO citation).
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk.
While we are not aware of evidence demonstrating the superiority of fidaxomicin
for reducing severe CDI outcomes, the reduced risk of recurrence

NNT x NNS = NNB

NNT for fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is vancomycin
because it's cheaper than fidaxomicin, even though IDSA recommends fidaxomicin.

rough estimate of costs.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.

@@

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @stuart_clinical_2021.

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have full datasets which use all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were pre-processed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. 
During pre-processing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
No features had missing values and no features were perfectly correlated.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.

@@

# Acknowledgements

TODO

@@

>>>>

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits.

@@

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
Similarly, the number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit from the treatment.
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
Thus the NNB pairs model performance with treatment effectiveness to estimate
the benefit of using predictive models in clinical practice.

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
rate of sustained resolution of CDI and lower rate of recurrence
[@stuart_clinical_2021].
The NNTs of fidaxomicin for sustained resolution and prevention of recurrence are
each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
However, fidaxomicin is considerably more expensive than vancomycin.
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk, while prescribing
vancomycin to low-risk patients.
If we assume that the superior efficacy of fidaxomicin for sustained resolution
and reduced recurrence also translates to reducing the risk of severe outcomes,
we can pair the NNT of fidaxomicin with the NNS of OTU-based prediction models
to estimate the NNB.

We determined the 95th percentile of risk for each prediction model and computed
the confusion matrix and several performance metrics including the NNS at this
threshold (@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than the prior curated model, and it is the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

TODO NNB
<!--
rough estimate of costs.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.
-->

@@

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @stuart_clinical_2021, e.g. @jiang_budget_2022.

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have full datasets which use all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were pre-processed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. 
During pre-processing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
having `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats` features
depending on the severity definition.
No features had missing values and no features were perfectly correlated.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.

@@

<!--
# Acknowledgements

TODO funding & author contributions
-->

@@

# Tables
{{< pagebreak >}}
